---
title: "Evaluation of a Rapid Diagnostic test (RDT) to Diagnose a Parasite-borne Disease"
author: "Obaroene Ogidiagba"
date: "2023-05-23"
output:
    html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preprocessing

```{r Imorting the librariers}
library(dplyr)    # For data manipulation
library(ggplot2)  # For data visualization
library(tableone) # For creating descriptive statistics
library(caTools)
library(caret)
library(pROC)
```

```{r r Import the data and inspect data}
data <- read.csv("dataset_to_analyse.csv")
# Data summary
str(data)
```

```{r Convert integer variables to factor}
data <- data %>%
  mutate(across(c(
    GENDER, DISEASE_SYMPT, INDEX_DONE, INDEX_LOT_N, INDEX_RES_CTRL,
    INDEX_RES_SPEC_1, INDEX_RES_SPEC_2, INDEX_POS, REF_RDT_DONE,
    REF_RDT_RES_CTRL, REF_RDT_RES_SPEC_1, REF_RDT_RES_SPEC_2,
    REF_RDT_POS, REF_MICRO_POS, REF_MICRO_SPEC
  ), as.factor))

# Data summary
str(data)
```

```{r Descriptive statistics for the dataset}
#Creating a list of all variables
myVars <- c("Participant.ID", "site", "DOB", "AGE", "GENDER", "DISEASE_SYMPT",
            "INDEX_DONE", "INDEX_LOT_N", "INDEX_RES_CTRL",
            "INDEX_RES_SPEC_1", "INDEX_RES_SPEC_2", "INDEX_POS",
            "REF_RDT_DONE", "REF_RDT_RES_CTRL", "REF_RDT_RES_SPEC_1",
            "REF_RDT_RES_SPEC_2", "REF_RDT_POS", "REF_MICRO_POS",
            "REF_MICRO_SPEC", "BIOMARKER_LEV")

#Creating a list of all categorical variables
catVars <- c("GENDER", "DISEASE_SYMPT",
             "INDEX_DONE", "INDEX_LOT_N", "INDEX_RES_CTRL",
             "INDEX_RES_SPEC_1", "INDEX_RES_SPEC_2", "INDEX_POS",
             "REF_RDT_DONE", "REF_RDT_RES_CTRL", "REF_RDT_RES_SPEC_1",
             "REF_RDT_RES_SPEC_2", "REF_RDT_POS", "REF_MICRO_POS",
             "REF_MICRO_SPEC")

# Creating a descriptive statistics table of categorical variables
Des_Table <- CreateTableOne(vars = myVars, data = data, factorVars = catVars)

# Print the dataframe
print(Des_Table, showAllLevels = TRUE)

#Creating a summary of the categorical and continous variables with percentage of missingness
summary(Des_Table$ContTable)  # Continous variables
summary(Des_Table$CatTable)   # Categorical variables

#Drop REF_MICRO_SPEC as it has 98.8% of data N/A
#N:B REF_MICRO_POS has 85.4% data missing, Age 37% and DOB with 63% missing
data <- data %>%
  select(-REF_MICRO_SPEC)
```

```{r Exploring continous variables}
# Histogram for distribution for Biomarker level
ggplot(data, aes(x = BIOMARKER_LEV)) +
  geom_histogram(fill = "lightblue", color = "black") +
  labs(title = "Distribution of Biomarker Levels", x = "Biomarker Level", y = "Frequency")

#Boxplot to check for outliers for Biomarker level
ggplot(data, aes(y = BIOMARKER_LEV)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Biomarkers outliers", y = "Frequency")

# Histogram for distribution for AGE
ggplot(data, aes(x = AGE)) +
  geom_histogram(fill = "lightblue", color = "black") +
  labs(title = "Distribution of Age", x = "Biomarker Level", y = "Frequency")

#Boxplot to check for outliers for AGE
ggplot(data, aes(y = AGE)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Age Outliers", y = "Frequency")

#summary of Age
summary(data$AGE)   #Maximum age is 243

# Remove observations with age > 125 from the data
data <- data[data$AGE <= 125, ]
summary(data$AGE)    #Maximum age now 85

# Histogram for distribution for AGE after removing outliers
ggplot(data, aes(x = AGE)) +
  geom_histogram(fill = "lightblue", color = "black") +
  labs(title = "Distribution of Age with max of 85", x = "Biomarker Level", y = "Frequency")

#Boxplot to check for outliers for AGE after removing outliers
ggplot(data, aes(y = AGE)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Boxplot of Age max of 85", y = "Frequency")

```

## Performance of the novel index RDT vs the microscopy and the reference RDT

```{r Index RDT vs Reference Microscopy}
# Create contingency tables for microscopy and index RDT
table_index_micro <- table(data$INDEX_POS, data$REF_MICRO_POS)

# Calculate performance metrics for microscopy and index RDT
metrics_index_micro <- c(
  sensitivity = table_index_micro[2, 2] / sum(table_index_micro[, 2]),
  specificity = table_index_micro[1, 1] / sum(table_index_micro[, 1]),
  ppv = table_index_micro[2, 2] / sum(table_index_micro[2, ]),
  npv = table_index_micro[1, 1] / sum(table_index_micro[1, ]),
  accuracy = sum(diag(table_index_micro)) / sum(table_index_micro)
)

# Print the performance metrics for microscopy and index RDT
print("Index RDT vs. Microscopy:")
print(metrics_index_micro)

# Perform McNemar's test
mcnemar_test_index_micro <- mcnemar.test(table_index_micro)
p_value_index_micro <- mcnemar_test_index_micro$p.value

# Calculate the confidence intervals
n1 <- table_index_micro[1, 2] + table_index_micro[2, 2]
n2 <- table_index_micro[2, 1] + table_index_micro[1, 1]

# Calculate the lower and upper bounds of the confidence interval
lower_bound <- (table_index_micro[2, 1] - table_index_micro[1, 2]) / n1 - 1.96 * sqrt((table_index_micro[2, 1] + table_index_micro[1, 2]) / n1^2)
upper_bound <- (table_index_micro[2, 1] - table_index_micro[1, 2]) / n1 + 1.96 * sqrt((table_index_micro[2, 1] + table_index_micro[1, 2]) / n1^2)

# Print the McNemar's test result and confidence intervals
cat("McNemar's test p-value:", p_value_index_micro, "\n")
cat("Confidence Intervals:", lower_bound, upper_bound, "\n\n")


```

```{r Index RDT vs Reference RDT}
# Create contingency table for index RDT and Reference RDT
table_index_ref <- table(data$INDEX_POS, data$REF_RDT_POS)

# Calculate performance metrics for index RDT and reference RDT
metrics_index_ref <- c(
  sensitivity = table_index_ref[2, 2] / sum(table_index_ref[, 2]),
  specificity = table_index_ref[1, 1] / sum(table_index_ref[, 1]),
  ppv = table_index_ref[2, 2] / sum(table_index_ref[2, ]),
  npv = table_index_ref[1, 1] / sum(table_index_ref[1, ]),
  accuracy = sum(diag(table_index_ref)) / sum(table_index_ref)
)

# Print the performance metrics
print("Index RDT vs. Reference RDT:")
print(metrics_index_ref)

# Perform McNemar's test
mcnemar_test_index_ref <- mcnemar.test(table_index_ref)
p_value_index_ref <- mcnemar_test_index_ref$p.value

# Calculate the confidence intervals
n1 <- table_index_ref[1, 2] + table_index_ref[2, 2]
n2 <- table_index_ref[2, 1] + table_index_ref[1, 1]

# Calculate the lower and upper bounds of the confidence interval
lower_bound <- (table_index_ref[2, 1] - table_index_ref[1, 2]) / n1 - 1.96 * sqrt((table_index_ref[2, 1] + table_index_ref[1, 2]) / n1^2)
upper_bound <- (table_index_ref[2, 1] - table_index_ref[1, 2]) / n1 + 1.96 * sqrt((table_index_ref[2, 1] + table_index_ref[1, 2]) / n1^2)

# Print the McNemar's test result and confidence intervals
cat("McNemar's test p-value:", p_value_index_ref, "\n")
cat("Confidence Intervals:", lower_bound, upper_bound, "\n\n")

```

## Patient status prediction using the biomaker level

#### Missing data will be handled using a complete case analysis. After appling the little MCAR test to determine if the data was MCAR or MAR, the test showed a statistically significant result. The choice of complete case analysis can potentially introduce some bias into the study but this was chosen due to the limited insight into the sampling and data collection for this study

```{r Data Exploration and Feature Selection }
# Create a new dataframe with selected variables
complete_data_ref_micro <- data[, c("BIOMARKER_LEV", "REF_MICRO_POS")]

complete_Table <- CreateTableOne(vars = c("BIOMARKER_LEV", "REF_MICRO_POS"),data = complete_data_ref_micro, factorVars = c("REF_MICRO_POS"))
print(complete_Table, showAllLevels = TRUE)

# Remove rows with missing data
complete_data_ref_micro <- complete_data_ref_micro[complete.cases(complete_data_ref_micro), ]

#Histogram to check for the distribution
ggplot(complete_data_ref_micro, aes(x = BIOMARKER_LEV)) +
  geom_histogram(fill = "lightblue", color = "black") +
  labs(title = "Distribution of Biomarker Levels", x = "Biomarker Level", y = "Frequency")

# Box plot of biomarker level by test result for reference microscopy
ggplot(complete_data_ref_micro, aes(x = REF_MICRO_POS, y = BIOMARKER_LEV, fill = REF_MICRO_POS)) +
  geom_boxplot() +
  labs(title = "Comparison of Biomarker Levels between Groups", x = "Microscopy Results", y = "Biomarker Level") +
  scale_fill_manual(values = c("lightblue", "lightgreen"), labels = c("Negative", "Positive"))

```

```{r Mann Whitney U Test}
# Split the dataset into two groups based on microscopy results
positive_group <- filter(complete_data_ref_micro, REF_MICRO_POS == 1)
negative_group <- filter(complete_data_ref_micro, REF_MICRO_POS == 0)

# Compare biomarker levels between the two groups by performing a Mann-Whitney U test
# Mann witney U preffered as the distribution of Biomarker levels were non-normal
mannwhitney_test_result <- wilcox.test(positive_group$BIOMARKER_LEV, negative_group$BIOMARKER_LEV)

# Print the Mann-Whitney U test results
cat("Mann-Whitney U Test Results:\n")
cat("p-value:", mannwhitney_test_result$p.value, "\n")
cat("Test statistic:", mannwhitney_test_result$statistic, "\n")

```

```{r Logistic Regression}
# Set the seed for reproducibility
set.seed(123)

# Splitting data into training and test set
split <- sample.split(complete_data_ref_micro$REF_MICRO_POS, SplitRatio = 0.8)
train_data <- complete_data_ref_micro[split, ]
test_data <- complete_data_ref_micro[!split, ]

# Logistic regression using training set
model <- glm(REF_MICRO_POS ~ BIOMARKER_LEV, data = train_data, family = "binomial")

# Summary of the regression
print(summary(model))

#calculate confidence interval at 95% confidence level
confint(model, 'BIOMARKER_LEV', level=0.95)

# Predicting the test set
predicted_values <- predict(model, newdata = test_data, type = "response")
predicted_classes <- ifelse(predicted_values > 0.5, "1", "0")

# Creating factors with the same levels
predicted_classes <- factor(predicted_classes, levels = c("1", "0"))
reference <- factor(test_data$REF_MICRO_POS, levels = c("1", "0"))

# Creating confusion matrix
confusion_matrix <- confusionMatrix(data = predicted_classes, reference = reference)

# Printing confusion matrix and statistics
print(confusion_matrix)

# Poor predictive ability or sensitivity maybe due to class imbalance
#Interpretation: he estimated coefficient for the biomarker level is 0.020090.It indicates that for a one-unit increase in the biomarker level, the log-odds of the response variable being 1 (positive) increases by 0.020090.  
#both the intercept and the biomarker level coefficient are highly significant (p < 0.001), indicating that they have a significant impact on the response variable.
  
# ROC curve and AUC
roc_obj <- roc(response = reference, predictor = predicted_values)
plot(roc_obj, main = "ROC Curve")
auc <- auc(roc_obj)
print(paste("AUC:", auc))  

#n AUC of 0.766304347826087 suggests that the logistic regression model has a moderately good discriminatory ability. It can differentiate between positive and negative cases with a probability of 0.766304347826087, which is better than random guessing but not perfect. 

# Residual analysis
residuals <- residuals(model)

# Plotting residuals
plot(residuals, main = "Residual Plot", xlab = "Observation", ylab = "Residuals")

# Normality test of residuals
shapiro.test(residuals)

# The residuals of the model are fitted around the 0 value of the plot.

```

##Sub-group analysis

```{r Performance of index test by symptoms}
symptoms_vs_index <- data.frame(table(data$DISEASE_SYMPT, data$INDEX_POS))
names(symptoms_vs_index) <- c("DISEASE_SYMPT", "INDEX_POS", "Count")

# Create the bar and legend
ggplot(data = symptoms_vs_index, aes(x = DISEASE_SYMPT, y = Count, fill = factor(INDEX_POS))) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("aquamarine4", "darkgrey"),
                    labels = c("Negative", "Positive"),
                    name = "Index Pos") +
  labs(title = "Performance of index test by symptoms",
       x = "Disease Symptoms", y = "Count") +
  theme_minimal()

#---Chi Squared test
# Chi-squared test of difference
chisq_result <- chisq.test(table(data$DISEASE_SYMPT, data$INDEX_POS))
print(chisq_result)
#The result suggests that there is no significant relationship or association between the disease symptoms and the index test positivity.

```

#### The logistic regression was also conducted by including potential confounding varaibles such as symptoms presence, age, gender. The results were not significantly different

```{r Performance of index test by Age}
# Filter out N/A values in INDEX_POS and Age
filtered_data_age_indexRDT <- data[complete.cases(data$INDEX_POS, data$AGE),]

# Box plot of biomarker level by test result for reference microscopy
ggplot(filtered_data_age_indexRDT, aes(x = INDEX_POS, y = BIOMARKER_LEV, fill = INDEX_POS)) +
  geom_boxplot() +
  labs(title = "Performance of index test by Age", x = "INDEX_POS", y = "Age") +
  scale_fill_manual(values = c("lightblue", "lightgreen"), labels = c("No", "Yes"))

#---- t-test of difference
# Perform independent samples t-test with missing values removed
t_result <- t.test(data$AGE ~ data$INDEX_POS, na.action = na.omit)

# Print the t-test result
print(t_result)
# The p-value (0.6601) is greater than the significance level (commonly set at 0.05), suggesting that there is no significant difference in the means of the two groups.

```

```{r Performance of index test by gender}
gender_vs_index <- data.frame(table(data$GENDER, data$INDEX_POS))
names(gender_vs_index) <- c("GENDER", "INDEX_POS", "Count")

# Create the bar and legend
ggplot(data = gender_vs_index, aes(x = GENDER, y = Count, fill = factor(INDEX_POS))) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("aquamarine4", "darkgrey"),
                    labels = c("Negative", "Positive"),
                    name = "Index Pos") +
  labs(title = "Performance of index test by Gender",
       x = "Gender", y = "Count") +
  theme_minimal()

# Chi-squared test of difference
chisq_result <- chisq.test(table(data$GENDER, data$INDEX_POS))
print(chisq_result)
#The result indicates that there is a statistically significant difference between the index test and gender. The low p-value suggests that the association between the variables is unlikely to be due to random chance alone.

```
